---
title: "Were those League Trends a Fluke?"
author: "Justin Pretorius"
date: "last-modified"
format:
  html:
    toc: true
    toc-depth: 2
    theme: sandstone
    highlight-style: tango
    number-sections: false
    df-print: paged
    code-fold: true
    code-tools: true
    embed-resources: true
    css: css/styles.css
execute: 
  echo: false
  warning: false
  message: false
editor:
  render-on-save: false
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: Setup
#| include: false

rm(list = ls())
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = 'hold',
  out.width = '100%',
  fig.width = 10,
  fig.asp = 0.7,
  fig.show = 'hold'
)
```

```{r}
#| label: Loading_Dependencies
#| include: false

packages = c(
  'quarto',
  'htmltools',
  'htmlwidgets',
  'reactable', 
  'reactablefmtr',
  'base64enc',
  'tidyverse',
  'plotly',
  'here',
  'broom',
  'gt',
  'purrr'
)

check_packages = function(packages, repos = '[https://cloud.r-project.org](https://cloud.r-project.org)') {
  invisible(
    lapply(packages, function(p) {
      if (!requireNamespace(p, quietly = TRUE)) {
        message(paste('Package', p, 'not found. Attempting to install...'))
        install.packages(p, repos = repos)
      }
      suppressPackageStartupMessages(
        library(p, character.only = TRUE)
      )
    })
  )
}
check_packages(packages)

helpers_file = here('kickers', 'helpers', 'helpers.R')
source(helpers_file)
```

```{r}
#| label: Loading Data

pbp_pp_file = here('kickers', 'data', 'pp', 'pbp', 'pbp_pp.rdata')
if (file.exists(pbp_pp_file)) {
  load(pbp_pp_file)
} 

kicker_trends_file = here(
  'kickers',
  'data',
  'analytic',
  'kicker_trends',
  'kicker_trends.rdata'
)
if (file.exists(kicker_trends_file)) {
  load(kicker_trends_file)
}

league_trends_file = here(
  'kickers',
  'data',
  'analytic',
  'league_trends',
  'league_trends.rdata'
)
if (file.exists(league_trends_file)) {
  load(league_trends_file)
} 

stadiums_file = here(
  'kickers',
  'data',
  'stadiums',
  'stadiums.rdata'
)
if (file.exists(stadiums_file)) {
  load(stadiums_file)
} 
```

## Review & Refresh
In my first blog (published centuries ago...), I shared some league-wide trends that I found interesting. These trends covered the increased athleticism and kicking performance of starting NFL kickers.

In short...
<ul>
<li> They are now taller and weigh less </li>
  <ul>
    <li> Don't make them like they used to </li>
    <li> Take a moment to appreciate the greatness that was Janikowski </li>
  </ul>
<li> Less field goals (FGs) are being attempted at distances less than 30 yards out </li>
<li> More FGs are being attempted at distances greater than or equal to 50 yards out </li>
<li> Accuracy has increased across the board, regardless the distance </li>
</ul>

What I didn't do when exploring these trends is control for a bunch of obvious stuff, like stadium roof type, weather, etc. Why? I wanted to leave you all wanting more.

With all that said, are you ready for this???

LET'S GET TO IT!!!

**Quick Note**
All charts are interactive. Don't be afraid to play around with them.

## NFL Starter Coverage
So my first blog focused solely on starting kickers. What this means is I carefully determined the starting kickers for each season from 2001 to 2025. How did I do this? If you want to know, I will happily share. To put it simply, I used some machine learning.

More importantly, how well did my approach work?

Looking at the table below, it should apparent that my data accounts for most of the field goals attempted per season. There are some years that are slightly concerning, like 2017 and 2018, but I'm not losing any sleep over it, and neither should you. These years are largely the result of high inconsistency at the starting kicker position. Like, what is going on with 2017? I am so tempted to launch an investigation into that year alone. For starters to account for only 46% of the kickers that year is crazy.

What you need to take away from this is here on out I will make it clear if I am referring all kickers or only starters. This distinction is important. Keep an eye out for it.

```{r}
#| label: NFL Starter Saturation

league_totals = pbp_pp %>%
  group_by(season) %>%
  summarise(
    total_kickers = n_distinct(kicker_player_id),
    total_fga = n(), 
    .groups = 'drop'
  )

starters = kicker_trends %>%
  group_by(season) %>%
  summarise(
    starters = n_distinct(kicker_player_id),
    starter_fga = sum(total_fga),
    .groups = 'drop'
  )

starter_coverage = league_totals %>%
  inner_join(y = starters, by = join_by(season)) %>%
  mutate(
    starter_coverage = starters / total_kickers,
    fga_coverage = starter_fga / total_fga
  ) %>%
  dplyr::select(
    season,
    starters, total_kickers, starter_coverage, 
    starter_fga, total_fga, fga_coverage
  ) %>%
  arrange(season)

col_defs = list(
  season = colDef(name = 'Season', align = 'left'),
  starters = colDef(name = 'Starters', align = 'left'),
  total_kickers = colDef(name = 'Total Kickers', align = 'left'),
  starter_coverage = colDef(
    name = 'Starter Coverage',
    align = 'left',
    format = colFormat(percent = TRUE, digits = 1)
  ),
  starter_fga = colDef(name = 'Starter FGA', align = 'left'),
  total_fga = colDef(name = 'Total FGA', align = 'left'),
  fga_coverage = colDef(
    name = 'FGA Coverage',
    align = 'left',
    format = colFormat(percent = TRUE, digits = 1)
  )
)
download_button = download_link(
  data = starter_coverage,
  filename = 'starter_coverage.csv'
)
strtr_cvg_table = reactable(
  data = starter_coverage,
  columns = col_defs,
  sortable = TRUE,
  filterable = TRUE,
  searchable = TRUE,
  pagination = TRUE,
  defaultPageSize = 8,
  showPageSizeOptions = TRUE,
  striped = TRUE,
  highlight = TRUE
)
tagList(
  download_button,
  strtr_cvg_table,
  tags$div(
    style = paste(
      'font-size: 0.85rem;',
      'color: #666;',
      'margin-top: 10px;',
      'font-style: italic;'
    ),
    '* FGA stands for Field Goals Attempted.'
  )
)
```

## The Obvious
Did you watch SpongeBob as a kid? Do you kids watch SpongeBob? That freakin' sponge will never go away. He's like the damn gum on the bottom of your shoe (haha). There was an episode where he became a hermit and feared going outside. Sandy eventually convinced him to come outdoors, but shortly after, he was mangled by a gorilla.

The point is, chilling out inside is undeniably nice sometimes, especially when winter is here. It's not far fetched to think that kickers feel the same way. Let's take a look at the trends in FG attempts, FG%, and the league median FG distance over the years when kicking inside compared to outside. <b>These trends were calculated using all NFL kickers.</b> 

The charts tell us three key points.
<ul>
 <li> There have been more FG attempts outside than inside, but the number of FG attempts inside is slowly catching up (more on this later) </li>
 <li> For most years, FG% tends to be better inside than outside </li>
 <li> The league median kick distance inside is usually greater than that outside - makes sense </li>
</ul>

```{r}
#| label: Indoors vs. Outdoors

in_vs_out = pbp_pp %>%
  dplyr::select(
    season, roof_clean, kick_distance, fg_result_clean
  ) %>%
  mutate(roof_clean = as.factor(roof_clean)) %>%
  group_by(season, roof_clean) %>%
  summarise(
    med_kck_dst = median(kick_distance, na.rm = TRUE),
    fg_att = n(),
    fg_made = sum(fg_result_clean == 'made'),
    fg_pct = fg_made/fg_att,
    .groups = 'drop'
  )

x_col = 'season'
y_cols = c('fg_pct', 'fg_att', 'med_kck_dst')
group_col = 'roof_clean'
grouped_line_plot(in_vs_out, x_col, y_cols, group_col)
```

Let's focus on FG attempts. Do our eyes deceive us? Are they really catching up to the number of FGs attempted outside? Yeah, they are - the stats clearly support this narrative. 

```{r}
#| label: Total FGA by Season

indoor_fga = in_vs_out %>%
  filter(roof_clean == 'indoors') %>%
  pull(fg_att)

outdoor_fga = in_vs_out %>%
  filter(roof_clean == 'outdoors') %>%
  pull(fg_att)

total_fga = indoor_fga + outdoor_fga
years = 2001:2025
tot_fga_tnd_tst = prop.trend.test(x = indoor_fga, n = total_fga, score = years)

tot_fga_tnd_tst %>%
  tidy() %>%
  gt() %>%
  tab_header(
    title = md('<b>Indoor vs. Outdoor FG Attempts (2001–2025)</b>'),
    subtitle = 'Chi-squared Test for Trend in Proportions'
  ) %>%
  cols_hide(columns = method) %>%
  cols_label(
    statistic = 'Chi-squared',
    p.value = 'p-value',
    parameter = 'df'
  ) %>%
  fmt_number(columns = statistic, decimals = 2) %>%
  fmt_scientific(columns = p.value, rows = p.value < 0.001) %>%
  tab_options(table.width = pct(80))
```

This result raises another important question. Is the league transitioning toward more stadiums that are, or can potentially be, indoors? 

Oddly enough, the number of domes has shockingly decreased. BUT... there are now more stadiums with retractable roofs. Even though the total number of stadiums that are, or can potentially be, indoors has NOT increased, the number of stadiums with retractable roofs has. Again, the stats support this narrative, and it could be one reason why the number of FGs attempted inside has also increased. 

```{r}
#| label: Stradium Growth by Season

stadium_roofs = pbp_pp %>%
  dplyr::select(season, stadium_id) %>%
  left_join(y = stadiums %>% 
              dplyr::select(stadium_id, roof_type),
            by = join_by(stadium_id)) %>%
  distinct() %>%
  group_by(season, roof_type) %>%
  summarise(count = n(), .groups = 'drop')

x_col = 'season'
y_cols = c('count')
group_col = 'roof_type'
grouped_line_plot(stadium_roofs, x_col, y_cols, group_col)
```

```{r}
#| label: Indoor Stadium Growth by Season

dome = stadium_roofs %>%
  filter(roof_type == 'Dome') %>%
  pull(count)

retractable = stadium_roofs %>%
  filter(roof_type == 'Retractable') %>%
  pull(count)

#' There were no retractable stadium roofs in 2001
#' Adding a zero equalizes the vectors in the addition step below
retractable = c(0, retractable)

indoor = dome + retractable

outdoor = stadium_roofs %>%
  filter(roof_type == 'Outdoors') %>%
  pull(count)

total_stdms = dome + retractable + outdoor
years = 2001:2025

indoor_grwth_tst = prop.trend.test(x = indoor, n = total_stdms, score = years)
indoor_grwth_tst %>%
  tidy() %>%
  gt() %>%
  tab_header(
    title = md('<b>Indoor Stadium Growth (2001–2025)</b>'),
    subtitle = 'Chi-squared Test for Trend in Proportions',
  ) %>%
  cols_hide(columns = method) %>%
  cols_label(
    statistic = 'Chi-squared',
    p.value = 'p-value',
    parameter = 'df'
  ) %>%
  fmt_number(columns = statistic, decimals = 2) %>%
  fmt_scientific(columns = p.value, rows = p.value < 0.001) %>%
  tab_options(table.width = pct(80))
```

```{r}
#| label: Retractable Stadium Growth by Season

rtrct_grwth_tst = prop.trend.test(
  x = retractable, n = total_stdms, score = years
)
rtrct_grwth_tst %>%
  tidy() %>%
  gt() %>%
  tab_header(
    title = md('<b>Retractable Roof Stadium Growth (2001–2025)</b>'),
    subtitle = 'Chi-squared Test for Trend in Proportions',
  ) %>%
  cols_hide(columns = method) %>%
  cols_label(
    statistic = 'Chi-squared',
    p.value = 'p-value',
    parameter = 'df'
  ) %>%
  fmt_number(columns = statistic, decimals = 2) %>%
  fmt_scientific(columns = p.value, rows = p.value < 0.001) %>%
  tab_options(table.width = pct(80))
```

What does this trend mean for kicking performance? Does the stadium roof type really matter? Are the distance and success of any given kick actually impacted by whether it's attempted inside or outside?

According to the stats, FG kicks attempted inside have a higher conversion rate and are roughly 1 yard longer than those outside. Cool, so they both are. I guess being an indoor hermit ain't that bad. 

```{r}
#| label: FG Result by Roof Type - Chi Square

chisq_data = pbp_pp %>%
  dplyr::select(fg_result_clean, roof_clean) %>%
  mutate(
    fg_result_clean = as.factor(fg_result_clean),
    roof_clean = as.factor(roof_clean)
  )
chisq_tbl = table(chisq_data$roof_clean, chisq_data$fg_result_clean)
chisq_test = chisq.test(chisq_tbl)

chisq_test %>%
  tidy() %>%
  gt() %>%
  tab_header(
    title = md('<b>FG Result by Roof Type (2001–2025)</b>'),
    subtitle = 'Chi-squared Test of Independence',
  ) %>%
  cols_hide(columns = method) %>%
  cols_label(
    statistic = 'Chi-squared',
    p.value = 'p-value',
    parameter = 'df'
  ) %>%
  fmt_number(columns = statistic, decimals = 2) %>%
  fmt_scientific(columns = p.value, rows = p.value < 0.001) %>%
  tab_options(table.width = pct(80))
```

```{r}
#| label: Kick Distance by Roof Type - Mann-Whitney U

wilcox_data = pbp_pp %>%
  dplyr::select(kick_distance, roof_clean) %>%
  mutate(roof_clean = as.factor(roof_clean))

wilcox_tst = wilcox.test(
  formula = kick_distance ~ roof_clean, data = wilcox_data, 
  alternative = 'two.sided', conf.int = TRUE
)

wilcox_tst %>%
  tidy() %>%
  gt() %>%
  tab_header(
    title = md('<b>Kick Distance by Roof Type (2001–2025)</b>'),
    subtitle = 'Mann-Whitney U Test of Distributional Shift',
  ) %>%
  cols_hide(columns = c(method, alternative)) %>%
  cols_label(
    estimate = 'Locational Difference',
    statistic = 'W',
    p.value = 'p-value',
    conf.low = 'CI Lower',
    conf.high = 'CI Higher',
    alternative = 'Alternative'
  ) %>%
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 4) %>%
  fmt_scientific(columns = p.value, rows = p.value < 0.001) %>%
  tab_options(table.width = pct(80))
```

Did I mislead everyone by reporting a bunch of things in the first blog without controlling for the obvious? Long story short, nah, but it's more nuanced than I suggested. 

The FG% outside is increasing by 0.29% per season, whereas the FG% inside is increasing by 0.13% per season. Overall, there is an undoubted steady increase in FG% per season since 2001, but this rate of increase depends heavily on whether the kick is inside or outside. The faster rate of increase per season for outside kicks is actually remarkable when you think about it. 

Is it because of improved coaching, training, or athleticism? You can really go down some serious rabbit holes with that one. Getting back on track, let's look at the league median kick distance.

```{r}
#| label: FG% by Roof Type over Season

lm_data = pbp_pp %>%
  dplyr::select(season, roof_clean, kick_distance, fg_result_clean) %>%
  mutate(
    roof = factor(roof_clean, levels = c('outdoors', 'indoors')),
    season = season - min(season)
  ) %>%
  group_by(season, roof) %>%
  summarise(
    med_kck_dst = median(kick_distance, na.rm = TRUE),
    fg_att = n(),
    fg_made = sum(fg_result_clean == 'made'),
    fg_pct = fg_made/fg_att,
    .groups = 'drop'
  )

fgp_mod = lm(fg_pct ~ roof + season + roof*season, data = lm_data)
mod_stats = glance(fgp_mod)

t = qt(0.975, df = fgp_mod$df.residual)

footer_text = paste0(
  '**Model Fit:** R² = ', round(mod_stats$r.squared, 3), 
  ' | Adj. R² = ', round(mod_stats$adj.r.squared, 3),
  ' | F-statistic = ', round(mod_stats$statistic, 2),
  ' (df = ', mod_stats$df.residual, ')',
  ' | p = ', format.pval(mod_stats$p.value, digits = 3, eps = 0.001)
)

abbrvtn_text = paste('**CI stands for Confidence Interval**')

fgp_mod %>%
  tidy() %>%
  mutate(
    conf.low = estimate - (t * std.error),
    conf.high = estimate + (t * std.error),
    term = recode_values(
      term,
      '(Intercept)' ~ 'Baseline (Outside)',
      'roofindoors' ~ 'Inside Effect',
      'season' ~ 'Yearly Trend (Yds/Year)',
      'roofindoors:season' ~ 'Interaction'
    )
  ) %>%
  gt() %>%
  tab_header(
    title = md('<b>Impact of Being Inside on FG%</b>'),
    subtitle = 'Outdoor Accuracy is Getting Scary Good!'
  ) %>%
  tab_options(
    table.width = pct(100),
    container.width = pct(100),
    row_group.background.color = '#f4f4f4'
  ) %>%
  cols_label(
    term = 'Predictor',
    estimate = 'Estimate',
    std.error = 'Std. Error',
    conf.low = '95% CI Lower',
    conf.high = '95% CI Upper',
    p.value = 'p-value'
  ) %>%
  cols_hide(columns = statistic) %>%
  fmt_number(columns = c(estimate, std.error), decimals = 3) %>%
  fmt_scientific(columns = conf.low, decimals = 2) %>%
  fmt_scientific(columns = conf.high, decimals = 2) %>%
  fmt_scientific(columns = p.value, decimals = 2, rows = p.value < 0.05) %>%
  tab_source_note(source_note = md(footer_text)) %>%
  tab_source_note(source_note = md(abbrvtn_text))
```

```{r}
#| label: FG% by Roof Type over Season Chart

grid = expand.grid(
  season = seq(
    from = min(lm_data$season), to = max(lm_data$season), length.out = 100
  ),
  roof = unique(lm_data$roof)
)

t = qt(0.975, df = fgp_mod$df.residual)
preds = predict(fgp_mod, newdata = grid, se.fit = TRUE)
grid$y_fit = preds$fit
grid$y_low = preds$fit - (t * preds$se.fit)
grid$y_high = preds$fit + (t * preds$se.fit)

plot_ly() %>%
  add_markers(
    data = lm_data, x = ~season, y = ~fg_pct, color = ~roof,
    colors = c('#E69F00', '#56B4E9'),
    legendgroup = ~roof,
    customdata = ~roof,
    hovertemplate = paste0(
      'Season: %{x}<br>',
      'FG%: %{y:.2%}<br>',
      'Roof Type: %{customdata}',
      '<extra></extra>'
    )
  ) %>%
  add_ribbons(
    data = grid, x = ~season, ymin = ~y_low, ymax = ~y_high, 
    color = ~roof, opacity = 0.2, 
    legendgroup = ~roof, showlegend = FALSE,
    hoverinfo = 'none'
  ) %>%
  add_lines(
    data = grid, x = ~season, y = ~y_fit, color = ~roof,
    line = list(width = 3), 
    legendgroup = ~roof, showlegend = FALSE,
    hovertemplate = '<b>Trend:</b> %{y:.2%}<extra></extra>'
  ) %>%
  layout(
    title = 'FG% Convergence: Outside vs. Inside',
    yaxis = list(title = 'FG%', tickformat = '.0%'),
    xaxis = list(title = 'Season'),
    hovermode = 'none'
  )
```

Despite no interaction between the stadium roof type and the season, the results clearly indicate that the league median kick distance is increasing at a rate of 0.153 yards per season, regardless of whether a game is played indoors or outdoors. Moreover, indoor games have a consistent 0.86-yard advantage in median kick distance over outdoor games, but both settings are experiencing the same annual increase in median kick distance.

As of right now, this model is a sham. It couldn't predict the outcome of a coin flip, so let's take a look at the weather. Is Mother Nature really that scary?


```{r}
#| label: Median Kick Distance by Roof Type over Season

med_kd_null_mod = lm(med_kck_dst ~ season, data = lm_data)
med_kd_prtl_mod = lm(med_kck_dst ~ roof + season, data = lm_data)
med_kd_full_mod = lm(med_kck_dst ~ roof + season + roof*season, data = lm_data)

ftest_prtl = anova(med_kd_null_mod, med_kd_prtl_mod, test = 'F') %>%
  tidy() %>%
  filter(!is.na(p.value)) %>%
  pull(p.value)

ftest_full = anova(med_kd_prtl_mod, med_kd_full_mod, test = 'F') %>%
  tidy() %>%
  filter(!is.na(p.value)) %>%
  pull(p.value)

ftest_text = paste0(
  '**F-Tests:**<br>',
  '**Partial vs Null (Roof Effect):** p = ', format.pval(ftest_prtl, eps = 0.001),
  '<br>',
  '**Full vs Partial (Interaction):** p = ', format.pval(ftest_full, eps = 0.001)
)

models = list(
  'Null (Season Effect)' = med_kd_null_mod,
  'Partial (No Interaction)' = med_kd_prtl_mod,
  'Full (Interaction)' = med_kd_full_mod
)

all_models_tidy = map_dfr(models, tidy, .id = 'model_name') %>%
  mutate(
    term = case_match(
      term,
      '(Intercept)' ~ 'Baseline (Outside)',
      'roofindoors' ~ 'Inside Effect',
      'season' ~ 'Yearly Trend (Yds/Year)',
      'roofindoors:season' ~ 'Interaction',
      .default = term
    ),
    display_val = paste0(
      round(estimate, 3), 
      ' (p=', 
      format.pval(p.value, digits = 3, eps = 0.01),
      ')'
    )
  )

table_data = all_models_tidy %>%
  select(model_name, term, display_val) %>%
  pivot_wider(names_from = model_name, values_from = display_val) %>%
  replace(is.na(.), '-')

footnote_text = paste0(
  '**Note:**<br>',
  'Estimates shown with p-values in parentheses. ',
  "The high p-value for the 'Interaction'",
  'suggests it does not significantly improve the model.'
)
table_data %>%
  gt() %>%
  tab_header(
    title = md('<b>Model Comparison: League Median Kick Distance</b>'),
    subtitle = 'Evaluating the necessity of Roof Type and Interaction effects'
  ) %>%
  tab_style(
    style = cell_text(weight = 'bold'),
    locations = cells_column_labels()
  ) %>%
  cols_label(term = 'Predictor') %>%
  tab_footnote(footnote = md(footnote_text)) %>%
  tab_source_note(source_note = md(ftest_text)) %>%
  tab_options(table.width = pct(100), data_row.padding = px(10))
```

Before I get into anything, here's how I broke down the weather for each outdoor game. If a game had any of the following, it was deemed as having bad weather or not ideal kicking conditions.

<ul>
<li> Precipitation (rain and snow) </li>
<li> Extreme cold (temperatures colder than 96% of all games) </li>
<li> Extreme heat (temperatures hotter than 96% of all games) </li>
</ul>

Every other outdoor game should have at least decent weather. Since we're talking about NFL kickers, decent weather should be good enough. Right?

Looking at the chart below, it's clear my approach is conservative, as most games seem to have decent or good weather. After all, we are talking about humans, too. Safety comes first, folks.

It is interesting to see the decrease in bad weather games. Again, this could be due to the league trending toward more retractable stadiums. 

```{r}
#| label: Good vs Bad Weather

weather_data = pbp_pp %>%
  dplyr::select(season, game_id, weather_clean) %>%
  distinct() %>%
  mutate(weather = if_else(weather_clean == 'bad', weather_clean, 'good')) %>%
  group_by(season, weather) %>%
  summarise(count = n(), .groups = 'drop')

xcol = 'season'
ycol = c('count')
group_col = 'weather'
grouped_line_plot(weather_data, xcol, ycol, group_col)
```

How do FG% and the league median kick distance stack up?

It turns out Mother Nature is unbiased. All her wrath was directed toward FG%. Good weather games, unsurprisingly, tend to have a higher seasonal FG% compared to bad weather games. Moreover, there is seemingly zero effect of weather on the league's median kick distance - more on this later.

But before we jump to any conclusions, let's see what the stats have to say.

```{r}
#| label: The Impact of Weather

weather_impact = pbp_pp %>%
  dplyr::select(
    season, roof_clean, weather_clean, kick_distance, fg_result_clean
  ) %>%
  filter(roof_clean == 'outdoors') %>%
  mutate(
    weather = if_else(weather_clean == 'bad', weather_clean, 'good'),
    weather = factor(weather, levels = c('bad', 'good'))
  ) %>%
  group_by(season, weather) %>%
  summarise(
    med_kck_dst = median(kick_distance, na.rm = TRUE),
    fg_att = n(),
    fg_made = sum(fg_result_clean == 'made'),
    fg_pct = fg_made/fg_att,
    .groups = 'drop'
  )

x_col = 'season'
y_cols = c('fg_pct', 'fg_att', 'med_kck_dst')
group_col = 'weather'
grouped_line_plot(weather_impact, x_col, y_cols, group_col)
```

It turns out the seasonal increase in FG% is really robust! Let's go! 

If I am going to be technically correct... 

Even if the weather stayed the same from year to year, we'd still see FG% climb about 0.3% every season. On top of that, playing in good weather effectively gives kickers a 2.1% accuracy increase over games played in the elements. Interesting, but nothing groundbreaking to report here.   

```{r}
#| label: FG% of Outdoor Kicks
outdoor_kicks = pbp_pp %>%
  dplyr::select(
    season, weather_clean, roof_clean, kick_distance, fg_result_clean
  ) %>%
  dplyr::filter(roof_clean == 'outdoors') %>%
  mutate(
    weather = if_else(weather_clean == 'bad', weather_clean, 'good'),
    weather = factor(weather, levels = c('bad', 'good')),
    season = season - min(season)
  ) %>%
  group_by(season, weather) %>%
  summarise(
    med_kck_dst = median(kick_distance, na.rm = TRUE),
    fg_att = n(),
    fg_made = sum(fg_result_clean == 'made'),
    fg_pct = fg_made/fg_att,
    .groups = 'drop'
  )

fgp_null_mod = lm(fg_pct ~ season, data = outdoor_kicks)
fgp_prtl_mod = lm(fg_pct ~ weather + season, data = outdoor_kicks)
fgp_full_mod = lm(fg_pct ~ weather + season + weather*season, data = outdoor_kicks)

ftest_prtl = anova(fgp_null_mod, fgp_prtl_mod, test = 'F') %>%
  tidy() %>%
  filter(!is.na(p.value)) %>%
  pull(p.value)

ftest_full = anova(fgp_prtl_mod, fgp_full_mod, test = 'F') %>%
  tidy() %>%
  filter(!is.na(p.value)) %>%
  pull(p.value)

ftest_text = paste0(
  '**F-Tests:**<br>',
  '**Partial vs Null (Roof Effect):** p = ', format.pval(ftest_prtl, eps = 0.001),
  '<br>',
  '**Full vs Partial (Interaction):** p = ', format.pval(ftest_full, eps = 0.001)
)

models = list(
  'Null (Season Effect)' = fgp_null_mod,
  'Partial (No Interaction)' = fgp_prtl_mod,
  'Full (Interaction)' = fgp_full_mod
)

all_models_tidy = map_dfr(models, tidy, .id = 'model_name') %>%
  mutate(
    term = case_match(
      term,
      '(Intercept)' ~ 'Baseline (Outside)',
      'weathergood' ~ 'Good Weather Effect',
      'season' ~ 'Yearly Trend (Yds/Year)',
      'weathergood:season' ~ 'Interaction',
      .default = term
    ),
    display_val = paste0(
      round(estimate, 3), 
      ' (p=', 
      format.pval(p.value, digits = 3, eps = 0.01),
      ')'
    )
  )

table_data = all_models_tidy %>%
  select(model_name, term, display_val) %>%
  pivot_wider(names_from = model_name, values_from = display_val) %>%
  replace(is.na(.), '-')

footnote_text = paste0(
  '**Note:**<br>',
  'Estimates shown with p-values in parentheses. ',
  "The high p-value for the 'Interaction'",
  'suggests it does not significantly improve the model.'
)
table_data %>%
  gt() %>%
  tab_header(
    title = md('<b>Model Comparison: The Impact of Weather on FG%</b>'),
    subtitle = 'Mother Nature Means Business!'
  ) %>%
  tab_style(
    style = cell_text(weight = 'bold'),
    locations = cells_column_labels()
  ) %>%
  cols_label(term = 'Predictor') %>%
  tab_footnote(footnote = md(footnote_text)) %>%
  tab_source_note(source_note = md(ftest_text)) %>%
  tab_options(table.width = pct(100), data_row.padding = px(10))
```

Moving on to the league median kick distance, the stats suggest that our eyes did not deceive us. For only outdoor kicks, it is unfazed by the weather. I'm gonna guess that many of you think this result is underwhelming.

WRONG

This result couldn't be more exciting because it means we can include both these terms in the model without unwanted effects. If you're a stats nerd, then you know I am referring to the curse of multicollinearity.

Are you ready for what's next? ARE YOU READY????

From here on out, we are really going to start really shifting gears and picking up the pace. These blogs will focus on building a robust model to predict the success of any given FG kick. 

So far, we've analyzed the impact of the following on the seasonal trends in FG% and league median kick distance.

<ul>
<li> Season </li>
<li> Stadium Roof Type </li>
<li> Weather </li>
</ul>

I've carefully structured the model so that the impact of weather on any given FG kick is completely removed from the prediction for indoor kicks. If you are interested in how I did this, let me know! I'd be happy to explain.

```{r}
#| label: Median Kick Distance of Outdoor Kicks

mkd_null_mod = lm(med_kck_dst ~ season, data = outdoor_kicks)
mkd_prtl_mod = lm(med_kck_dst ~ weather + season, data = outdoor_kicks)
mkd_full_mod = lm(
  med_kck_dst ~ weather + season + weather*season, data = outdoor_kicks
)

ftest_prtl = anova(mkd_null_mod, mkd_prtl_mod, test = 'F') %>%
  tidy() %>%
  filter(!is.na(p.value)) %>%
  pull(p.value)

ftest_full = anova(mkd_prtl_mod, mkd_full_mod, test = 'F') %>%
  tidy() %>%
  filter(!is.na(p.value)) %>%
  pull(p.value)

ftest_text = paste0(
  '**F-Tests:**<br>',
  '**Partial vs Null (Roof Effect):** p = ', format.pval(ftest_prtl, eps = 0.001),
  '<br>',
  '**Full vs Partial (Interaction):** p = ', format.pval(ftest_full, eps = 0.001)
)

models = list(
  'Null (Season Effect)' = mkd_null_mod,
  'Partial (No Interaction)' = mkd_prtl_mod,
  'Full (Interaction)' = mkd_full_mod
)

all_models_tidy = map_dfr(models, tidy, .id = 'model_name') %>%
  mutate(
    term = case_match(
      term,
      '(Intercept)' ~ 'Baseline (Outside)',
      'weathergood' ~ 'Good Weather Effect',
      'season' ~ 'Yearly Trend (Yds/Year)',
      'weathergood:season' ~ 'Interaction',
      .default = term
    ),
    display_val = paste0(
      round(estimate, 3), 
      ' (p=', 
      format.pval(p.value, digits = 3, eps = 0.01),
      ')'
    )
  )

table_data = all_models_tidy %>%
  select(model_name, term, display_val) %>%
  pivot_wider(names_from = model_name, values_from = display_val) %>%
  replace(is.na(.), '-')

footnote_text = paste0(
  '**Note:**<br>',
  'Estimates shown with p-values in parentheses. ',
  "The high p-value for the 'Good Weather Effect' and 'Interaction'",
  'suggests that neither term significantly improves the model.'
)
table_data %>%
  gt() %>%
  tab_header(
    title = md('<b>Model Comparison: League Median Kick Distance vs. Weather</b>'),
    subtitle = 'Rain or Shine - You Gonna Kick Far!'
  ) %>%
  tab_style(
    style = cell_text(weight = 'bold'),
    locations = cells_column_labels()
  ) %>%
  cols_label(term = 'Predictor') %>%
  tab_footnote(footnote = md(footnote_text)) %>%
  tab_source_note(source_note = md(ftest_text)) %>%
  tab_options(table.width = pct(100), data_row.padding = px(10))
```

## Time To Get Serious
We can clearly see that all terms in the model are statistically significant for predicting the success of any given FG kick. This is what you should take away from the results below.

<ul>
<li> For every 1-yard increase in kick distance, the odds of a make decrease by 10.1% </li>
<li> If we were to kick at the same distance every time, kicking outdoors in good weather reduces the odds of making a kick by 21.6% compared to kicking inside </li>
<li> If we are kicking outside, bad weather further reduces the odds of making a kick by an additional 16% compared to kicking in good weather </li>
</ul>

In other words, kicking outside in bad weather reduces the odds of making a kick by about 34% compared to kicking inside at the same distance. Hahaha holy crap that's a lot!!!

```{r}
#| label: xFG Model

xfg_data = pbp_pp %>% 
  dplyr::select(
    kicker_player_id, kicker_name_clean, 
    fg_result_clean, 
    kick_distance, roof_clean, weather_clean, season
  ) %>%
  mutate(
    season = season - min(season),
    fgr = if_else(fg_result_clean == 'made', 1, 0),
    roof = if_else(roof_clean == 'outdoors', 1, 0),
    weather = if_else(weather_clean == 'bad', 1, 0),
  )
xfg_mod = glm(
  formula = fgr ~ kick_distance + season + roof + roof:weather, 
  data = xfg_data,
  family = 'binomial'
)

xfg_tidy = tidy(xfg_mod, conf.int = TRUE) %>%
  mutate(
    across(c(estimate, conf.low, conf.high), exp),
    term = case_match(
      term,
      '(Intercept)' ~ 'Baseline (Inside)',
      'kick_distance' ~ 'Kick Distance (per Yard)',
      'season' ~ 'Yearly Trend',
      'roof' ~ 'Outside Effect (Good Weather)',
      'roof:weather' ~ 'Bad Weather Effect (Outside Kicks Only)'
    ),
    or_p = paste0(
      round(estimate, 3),
      ' (p=',
      format.pval(p.value, digits = 3, eps = 0.001),
      ')'
    ),
    ci_range = paste0(
      '[', round(conf.low, 3), ', ', round(conf.high, 3), ']'
    )
  )

xfg_tidy %>%
  dplyr::select(term, or_p, ci_range) %>%
  gt() %>%
  tab_header(
    title = md("<b>FG Probability: Environmental & Distance Model</b>"),
    subtitle = "Predicting the Odds of a Successful Conversion"
  ) %>%
  cols_label(
    term = "Predictor",
    or_p = "Odds Ratio (p-value)",
    ci_range = "95% CI"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_footnote(
    footnote = md(paste0(
      '**Note:** <br>',
      'Values are exponentiated Odds Ratios. ',
      'An OR < 1 indicates a decrease in the odds of success. ',
      'The *Bad Weather Effect* is conditional; it only applies to outside kicks.'
    ))
  ) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(10)
  )
```

If we focus on the 2025 season, we can even plot FG success probability curves for inside kicks, outside kicks in good weather, and outside kicks in bad weather. Look at the curve for outside kicks! What a drop off...

We can go even further and determine which kickers performed over the league average expectation in 2025. First, we gotta predict whether each kick should have been made or missed. Summing these probabilities over the 2025 season tells us how many kicks each kicker should have made. Subtracting this expectation from their actual total tells us who over- and underachieved.

Will Reichard is a freakin' beast... He accounted for four more FGs than an average league kicker if that kicker were given the same kicking conditions. His value to the Vikings is stupid. Twelve points can win a game in some match-ups. 

Cough... cough... Denver vs LV in Week 10...  

Check out the leaderboard here if you're interested -> [Leaderboard](https://justind.shinyapps.io/KickerLeaderboard/)

```{r}
#| label: Visualizing FG Probability Success Curves

min_kd = min(xfg_data$kick_distance)
max_kd = max(xfg_data$kick_distance)

plot_data = expand.grid(
  kick_distance = min_kd:max_kd,
  roof = c(0, 1),
  weather = c(0, 1),
  season = 24
) %>%
  filter(!(roof == 0 & weather == 1)) %>%
  mutate(
    prob = predict(xfg_mod, newdata = ., type = 'response'),
    scenario = case_when(
      roof == 0 ~ 'Indoors',
      roof == 1 & weather == 0 ~ 'Outdoors (Good Weather)',
      roof == 1 & weather == 1 ~ 'Outdoors (Bad Weather)'
    )
  )

plot_ly(
  plot_data, 
  x = ~kick_distance, 
  y = ~prob, 
  color = ~scenario, 
  colors = c('#2ecc71', '#e74c3c', '#3498db'),
  type = 'scatter', 
  mode = 'lines',
  hovertemplate = paste(
    '%{fullData.name}<br>',
    'Distance: %{x} yards<br>',
    'Prob: %{y:.1%}<extra></extra>'
  )
) %>%
  layout(
    title = 'FG Probability by Distance',
    xaxis = list(title = 'Kick Distance (Yards)'),
    yaxis = list(title = 'Success Probability', tickformat = '.0%'),
    hovermode = 'x unified',
    legend = list(orientation = 'h', y = -0.2)
  )
```

```{r}
#| label: Creating 2025 xFGOE Leaderboard

strtrs = kicker_trends %>%
  dplyr::select(season, kicker_name_clean, kicker_player_id) %>%
  mutate(season = season - 2001)

xfg_predictions = xfg_data %>%
  mutate(
    exp_fgr = predict(xfg_mod, type = 'response'),
    xfgoe = fgr - exp_fgr
  )

xfgoe_ldrbrd = xfg_predictions %>%
  inner_join(
    y = strtrs,
    by = join_by(season, kicker_player_id, kicker_name_clean)
  ) %>%
  filter(season == 24) %>%
  group_by(kicker_player_id, kicker_name_clean) %>% 
  dplyr::summarize(
    attempts = n(),
    makes = sum(fgr),
    exp_makes = round(sum(exp_fgr), 1),
    total_xfgoe = round(sum(xfgoe), 2),
    xfgoe_per_kck = round(mean(xfgoe), 3),
    .groups = 'drop'
  ) %>%
  arrange(desc(total_xfgoe))
```

```{r}
#| label: Visualizing 2025 xFGOE Leaderboard

plot_data = xfgoe_ldrbrd %>%
  slice_max(total_xfgoe, n = 20) %>%
  mutate(
    marker_color = if_else(total_xfgoe > 0, '#2ecc71', '#e74c3c'),
    kicker_label = reorder(kicker_name_clean, total_xfgoe),
    hover_text = paste(
      '<b>Kicker:</b>', kicker_name_clean, '<br>',
      '<b>Season: 2025</b><br>',
      '<b>xFGOE:</b>', total_xfgoe, ' kicks added<br>',
      '-----------------------<br>',
      '<b>Attempts:</b>', attempts, '<br>',
      '<b>Success:</b>', makes, 'Actual /', exp_makes, 'Expected'
    )
  )

plot_ly(plot_data) %>%
  add_segments(
    x = 0, xend = ~total_xfgoe, 
    y = ~kicker_label, yend = ~kicker_label,
    line = list(color = '#aaaaaa', width = 1),
    showlegend = FALSE, hoverinfo = 'none'
  ) %>%
  add_markers(
    x = ~total_xfgoe, 
    y = ~kicker_label,
    marker = list(
      size = 12, color = ~marker_color, 
      line = list(color = '#ffffff', width = 1)
    ),
    text = ~hover_text,
    hoverinfo = 'text',
    name = 'Performance'
  ) %>%
  layout(
    title = list(
      text = paste(
        '2025 Top 20 NFL Kickers<br>',
        'Based on Value Added (xFGOE)'
        ),
      y = 0.95
    ),
    xaxis = list(title = 'Total Kicks Added Above League Average'),
    yaxis = list(title = ""),
    margin = list(l = 150, r = 20, b = 60, t = 80),
    paper_bgcolor = '#f8f9fa', plot_bgcolor = '#f8f9fa'
  )
```

## Summary
Seems like a lot happened in this blog, so it'd be good to summarize everything before closing out.

<ul>
<li> Do starting kickers account for most of the FGs attempted in any given season? </li>
  <ul><li> Yeah... duh...</li></ul>
<li> Are really getting more accurate as the season march on? </li>
  <ul><li> Absolutely </li></ul>
<li> Are longer kicks a mirage? </li>
  <ul><li> No way, kick distance is increasing annually. </li></ul>
<li> How does kicking fare inside vs. outside? </li>
  <ul><li> Clearly SpongeBob knew something about the great outdoors. </li></ul>
<li> Why does every team seem to be getting a new stadium? </li>
  <ul><li> Cause they are, and they're all retractable. Sofi started a trend. </li></ul>
<li> How is Mother Nature treating outside FG attempts? </li>
  <ul>
    <li> She clearly has some issues with kickers. </li>
    <li> Their accuracy is taking quite a hit in bad weather games. </li>
    <li> But they're not afraid because they're still kicking from insane distances. </li>
  </ul>
<li> Who was the last year's Most Valuable Kicker (MVK)? </li>
  <ul>
    <li> Give Will Reichard a round of applause! </li>
    <li> He's currently atop the leaderboard, but it's gonna be fun to see how this changes as our model gets more accurate. </li>
  </ul>
</ul>

## Closing
If you thought that was cool, keep thinking. We are just getting started. In the next blog, I am going to take a look at how all the following impact FG% and the league median kick distance. Those that do will be added to our model to adjust the rankings. 

<ul>
<li> Time zone changes </li>
<li> Altitude </li>
<li> Back-to-back divisional games </li>
<li> Back-to-back games against opponents with records above .500 </li>
</ul>

## Social Media
* [GitHub](https://github.com/jdpretorius123)
* [X](https://x.com/jdpretorius_)
* [LinkedIn](https://www.linkedin.com/in/justin-p-996555172/)
* [Email](mailto:justin2025@gmail.com)